Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2024-11-29 paper/main.tex
Updates paper
Adds another plot

One of the common ways to guide this post-hoc modification of DAGs is to use
Conditional Independence (CI) based model testing \cite{Ankan2023}. As DAGs
imply CI statements that can be tested in the given data using statistical CI
tests, we can verify whether the implied CIs are consistent with the data or
not. If some of implied CIs are not valid in the given dataset, we can modify
the model accordingly. Figure~\ref{fig:ci_testing} shows an example of model
testing using CIs. However, this CI based testing has a few drawbacks: 1) CIs
are implied only by missing edges in the model, so we can not test for already
existing edges. 2) In high sample size scenarios, p-values can suggest a
significant association between variables even if they have a very weak
relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example,  in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

@@

\begin{figure}
	\centering
	\begin{tikzpicture}
		\begin{scope}[xshift=-2.5cm, yshift=0.7cm, scale=0.7]
			\tikzstyle{every node}=[align=center, inner sep=1pt]
			\node (sex) at (-0.7, -0.8) {Sex};
			\node (age) at (-0.3, 0) {Age};
			\node (ed) at (-1.8, -1.6) {Education};
			\node (occ) at (1.8, 0) {Occupation};
			\node (hrpw) at (1.3, -1.6) {Hours Per \\ Week};
			\node (income) at (-0.7, -2.8) {Income};

			\draw[->]  (age) to[bend right=20] (ed);
			\draw[->]  (sex) to (ed);
			\draw[->]  (age) to (hrpw);
			\draw[->]  (ed) to (hrpw);
			\draw[->]  (sex) to (hrpw);
			\draw[->]  (ed) to (income);
			\draw[->]  (hrpw) to (income);
			\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
			\draw[->]  (sex) to (income);
			\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
			\draw[->]  (sex) to (occ);	
		\end{scope}
		\begin{scope}[xshift=-3cm]
			\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
			\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
				\begin{minipage}{.2\textwidth}
					\begin{equation*}
						\begin{split}
							\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
							\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
							\textnormal{Age} &\ci \textnormal{Sex} \\
							\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
						\end{split}
					\end{equation*}
				\end{minipage}
				};
		\end{scope}
		\end{tikzpicture}
		\caption{\todo[inline]{Generate this figure using a few different CI tests and also show the unoriented edges in some of them}}
		\label{fig:intro}
\end{figure}

@@

\section{Connection to Score Based Methods}
\begin{itemize}
	\item The process of iteratively adding and removing edges in very similar to
		Greedy Equivalence Search (GES).
	\item But Because they make only local changes, can get stuck in local minimas.
	\item This approach looks for modifications globally.
\end{itemize}

\todo[inline]{Show an example where local changes can get stuck}

@@

\begin{figure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC and Hill-Climb Search algorithms against
		manually drawn DAGs using the assistance method. As PC and
		Hill-Climb Search return the Markov Equivalence Class (MEC), we
		use the best and worst scoring orientation of the MEC to get
		the range of values. The human drawn values are done for
	multiple accuracy values of the oracle. \todo[inline]{Replace Hill-Climb with GES and improve the plot}}
\end{figure}

>>>>

One of the common ways to guide this post-hoc modification of DAGs is to use
Conditional Independence (CI) based model testing \cite{Ankan2023}. As DAGs
imply CI statements that can be tested in the given data using statistical CI
tests, we can verify whether the implied CIs are consistent with the data or
not. If some of implied CIs are not valid in the given dataset, we can modify
the model accordingly. However, this CI based testing has a few drawbacks: 1) CIs
are implied only by missing edges in the model, so we can not test for already
existing edges. 2) In high sample size scenarios, p-values can suggest a
significant association between variables even if they have a very weak
relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example,  in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

@@

\begin{figure}
\centering
\begin{subfigure}{0.5 \textwidth}
	\begin{tikzpicture}
	\begin{scope}[xshift=-2.5cm, yshift=0.7cm]
		\tikzstyle{every node}=[align=center, inner sep=1pt]
		\node (sex) at (-0.7, -0.8) {Sex};
		\node (age) at (-0.3, 0) {Age};
		\node (ed) at (-1.8, -1.6) {Education};
		\node (occ) at (1.8, 0) {Occupation};
		\node (hrpw) at (1.3, -1.6) {Hours Per \\ Week};
		\node (income) at (-0.7, -2.8) {Income};

		\draw[->]  (age) to[bend right=20] (ed);
		\draw[->]  (sex) to (ed);
		\draw[->]  (age) to (hrpw);
		\draw[->]  (ed) to (hrpw);
		\draw[->]  (sex) to (hrpw);
		\draw[->]  (ed) to (income);
		\draw[->]  (hrpw) to (income);
		\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
		\draw[->]  (sex) to (income);
		\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
		\draw[->]  (sex) to (occ);	
	\end{scope}
	\end{tikzpicture}
\end{subfigure}
\begin{subfigure}{0.5 \textwidth}
	\begin{tikzpicture}
	\begin{scope}[xshift=-2.5cm, yshift=0.7cm]
		\tikzstyle{every node}=[align=center, inner sep=1pt]
		\node (wclass) at (0, 1) {WorkClass};
		\node (sex) at (0, -0.8) {Sex};
		\node (age) at (-0.3, 0) {Age};
		\node (ed) at (-2, -1.6) {Education};
		\node (occ) at (1.8, 0) {Occupation};
		\node (hrpw) at (1, -2) {Hours Per \\ Week};
		\node (income) at (-0.7, -2.3) {Income};
		\node (marital) at (3, 1) {Marital \\ Status};
		\node (reln) at (-3, 1) {Relationship};
		\node (race) at (-4, 2) {Race};
		\node (country) at (-4, 3) {Native \\ Country};

		\draw[] (ed) to (income);
		\draw[] (marital) to (reln);
		\draw[] (occ) to (sex);
		\draw[->] (reln) to (age);
		\draw[->] (reln) to (marital);
		\draw[] (reln) to (sex);
		\draw[->] (reln) to (hrpw);
		\draw[] (reln) to (income);
		\draw[->] (income) to (ed);
	\end{scope}
	\end{tikzpicture}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
	\begin{tikzpicture}
	\begin{scope}[xshift=-2.5cm, yshift=0.7cm]
		\tikzstyle{every node}=[align=center, inner sep=1pt]
		\node (wclass) at (0, 1) {WorkClass};
		\node (sex) at (0, -0.8) {Sex};
		\node (age) at (-0.3, 0) {Age};
		\node (ed) at (-2, -1.6) {Education};
		\node (occ) at (1.8, 0) {Occupation};
		\node (hrpw) at (1, -2) {Hours Per \\ Week};
		\node (income) at (-0.7, -2.3) {Income};
		\node (marital) at (3, 1) {Marital \\ Status};
		\node (reln) at (-3, 1) {Relationship};
		\node (race) at (-4, 2) {Race};
		\node (country) at (-4, 3) {Native \\ Country};

		\draw[]   (wclass) to (ed);
		\draw[]   (wclass) to (occ);
		\draw[]   (wclass) to (race);
		\draw[]   (wclass) to (income);
		\draw[]   (ed) to (occ);
		\draw[]   (ed) to (country);
		\draw[]   (ed) to (income);
		\draw[]   (marital) to (occ);
		\draw[]   (marital) to (reln);
		\draw[->] (marital) to (sex);
		\draw[]   (marital) to (hrpw);
		\draw[->] (occ) to (sex);
		\draw[]   (occ) to (hrpw);
		\draw[->] (reln) to (sex);
		\draw[]   (reln) to (hrpw);
		\draw[->] (income) to (occ);

	\end{scope}
	\end{tikzpicture}
\end{subfigure}

\caption{Markov Equivalence Class (MEC) learnt for Adult Income Dataset using (a) PC algorithm, (b) Hill-Climb Search with BIC score, (c) }
\label{fig:intro}
\end{figure}

			% \begin{scope}[xshift=-3cm]
			% 	\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
			% 	\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
			% 		\begin{minipage}{.2\textwidth}
			% 			\begin{equation*}
			% 				\begin{split}
			% 					\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
			% 					\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
			% 					\textnormal{Age} &\ci \textnormal{Sex} \\
			% 					\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
			% 				\end{split}
			% 			\end{equation*}
			% 		\end{minipage}
			% 		};
			% \end{scope}

@@

\section{Connection to Score Based Methods}
\begin{itemize}
	\item The process of iteratively adding and removing edges in very similar to
		Greedy Equivalence Search (GES).
	\item But not decomposible like scoring methods. 
	\item But Because they make only local changes, can get stuck in local minimas.
	\item Both approach looks for modifications globally.
	\item The sum of unexplained effect provides an absolute measure of model fit 
		unlike scoring methods.
\end{itemize}

@@

\begin{figure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/shd_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics{../code/fig3/sid_ribbon.pdf}
		\caption{}
	\end{subfigure}
	\caption{Comparison of PC and Hill-Climb Search algorithms against
		manually drawn DAGs using the assistance method. As PC and
		Hill-Climb Search return the Markov Equivalence Class (MEC), we
		use the best and worst scoring orientation of the MEC to get
		the range of values. The human drawn values are done for
		multiple accuracy values of the oracle.}
\end{figure}
